Stability and performance of coalition of prosumers through diversification in the Smart Grid

### 2015-08-03 Article.tex
update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Title
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Stability And Performance Of Virtual Aggregations Of Prosumers Through Diversification In The Smart Grid}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Authors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Nicolas Gensollen, Monique Becker, Vincent Gauthier and Michel Marot  \\
\IEEEauthorblockA{CNRS UMR 5157 SAMOVAR, \\
Telecom SudParis/Institut Mines Telecom\\
Email: \{nicolas.gensollen, vincent.gauthier, michel.marot, monique.becker\}@telecom-sudparis.eu}}

@@

%Designing stable power systems is a classical engineering challenge since blackout events can have catastrophic consequences. A recent approach introduced by complex systems theorists consists in abstracting the power grid as a graph where nodes represent loads, generators, or transformers, while edges stand for electrical lines \cite{Pagani2012} \cite{Rosas-Casals2007} \cite{Pahwa}. As graphs, power grids display some characteristics like clustering coefficients, degree distributions, or density for instance. These metrics enable scientists to get a better understanding of the structure of such complex systems \cite{Pagani2012}. It appears that power grids are particular graphs compared to well-known large networks like the internet, the world wide web, or social networks. That is, they exhibit different characteritics, suggesting that they may behave differently \cite{Rosas-Casals2007}. A common hypothesis for such a difference is that power grids are completely designed by a very small number of entities (electricity operators mainly), whereas other pre-cited graphs results from the interactions of billions of independant entities. Therefore, properties like power law degree distributions, prerential attachment, or hub oriented attacks do not necessarily apply with power grids.

%Actually, drawing conclusions based only on the study of the underlying graph structure may lead to unrealistic results \cite{Pahwa}. Power grids are ideed dynamical systems governed by physical laws. Concepts such as shortest path or centrality metrics have thus to be extended to take into account the particular nature of power grids. Besides, stability and resilience of such systems are much more complex than the simple connectivity of the underlying graph \cite{Pahwa}. It is well-known for instance that frequency synchronization in the grid is a necessary condition for stability. A common solution consists in modeling the grid as a network of coupled oscillators governed by a swing equation dynamic \cite{Filatrella2008}. It can then be shown that the topology of the network impacts directly the capacity of the oscillators to reach the synchronized state \cite{Dorfler2013}.

@@

In this paper, we focus on these "\textit{nodes}" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper). A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive as to satisfy both parts. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation append.

@@

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variability. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performances on the communication side \cite{Negeri2012}, since an aggregator can be considered as a single node by the level above it. On the other hand, it is well-known that diversification of the assets is a way of minimizing risks when constructing a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are obviously not equally stable or productive, which means that special attention should be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specify both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}).

@@

As explained in the section above, allowing entities using mainly renewables to inject power in the grid is a difficult challenge. Indeed, on the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the resource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Coupled with demand side management techniques implemented on the end users' smart meters, this could result in less production is wasted as well as smaller consumption peaks \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing is believed to be a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

@@

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the return distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{doi:10.1142/S0219024912500197}.

@@

\begin{equation}
\left\{ \begin{array}{lll}
		\mu_{ij} = \mu_{i} + \mu_{j} \\
		\sigma_{ij} = \sqrt{\sigma_{i}^{2} + \sigma_{j}^{2} + \rho_{ij} \sigma_{i} \sigma_{j} }
\end{array} \right.
\end{equation}

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants when $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this kind of situations, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalties to the grid. The probability that $ \{i,j\}$ is under-producing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P_{S}^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{S}^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

Where $ erf $ is the error function : $ erf(x) = \dfrac{2}{\sqrt{\pi}}\int_{0}^{x} e^{-t^{2}} dt $.

@@

\begin{equation}
gain(S) = \int_{t_{0}}^{t_{k}} \lambda P_{S}^{CRCT \star} dt = P_{S}^{CRCT \star} \int_{t_{0}}^{t_{k}} \lambda dt
\end{equation}

(since $ P_{S}^{CRCT \star} $ is supposed to be a constant rate over the contracted period). Using $ gain(S) $ directly as a utility function suffers a major drawback. It is indeed not a concave function of the coalitions' sizes, meaning that coalitions can grow as large as the number of agents allows it, without any counterbalance effects.

@@

\begin{equation}
\mathcal{U}(S) = \left\{ \begin{array}{lll}
							\dfrac{1}{|S|^{\alpha}} \dfrac{ P_{S}^{CRCT \star} }{P^{MAX}},\ if\ S\ is\ valid, \\
							0,\ if\ S\ is\ not\ valid
						 \end{array}
				  \right.
\end{equation}

Where parameter $ \alpha $ controls to what extent the size of a coalition impacts its utility, and $ P^{MAX} $ is a normalizing factor. $ P^{MAX} $ can be seen as the maximum production which can be injected in the grid.

@@

In order to get an estimator for $ \alpha $, we solve equation \ref{eq:derivative} in a Gaussian case (as in section \ref{sec:notations}). Furthermore, since considering all the possible interactions between agents is analytically intractable, we use here a mean approximation. Any quantity x that varies over the agent set is thus simplified in its mean value $ \bar{x} $. Solving equation \ref{eq:derivative} for $ \alpha $ in these conditions yields :

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not to study of coalitions stability against player defection which could be done through game theory, nor to redistribute the coalition's utility in terms of individual payoffs. But we aim to design $ \mathcal{U} $ as a measure of how good a given coalition is according to our criteria. In other terms does a given coalitions has a good production to risk ratio.

@@

%Let $ P^{MARKET} = \sum_{S \in CS} P_{S}^{CRCT} $ be the production on the market, where CS is the coalition structure formed $ CS = \{S_{1},...S_{N_{COAL}} \} $, and $ N_{COAL} $ is the number of coalitions in the structure. First of all, $ P^{MARKET} $ depends explicitely on $ N_{COAL} $ such that, forming more coalitions induces an increase of the production. However, for a finite pool of agents, we expect an optimal number of coalitions $ N_{COAL}^{\star} $ that maximizes the market production. The reason for this expectation is that, for a given number of agents, more coalitions causes necessarily less agents in the coalitions, and therefore less diversity. We expect thus a point where coalitions formed start to fail the market entrance, and deteriorate the market production.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Title
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Stability and performance of coalition of prosumers through diversification in the Smart Grid}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Authors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Nicolas Gensollen, Vincent Gauthier, Monique Becker and Michel Marot  \\
\IEEEauthorblockA{CNRS UMR 5157 SAMOVAR, \\
Telecom SudParis/Institut Mines Telecom\\
Email: \{nicolas.gensollen, vincent.gauthier, michel.marot, monique.becker\}@telecom-sudparis.eu}}

@@

@@

In this paper, we focus on these "\textit{nodes}" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper). A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation happen.

@@

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variability. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performance on the communication side \cite{Negeri2012}, since an aggregator can be considered as a single node by the level above it. On the other hand, it is well-known that diversification of the assets is a way of minimizing risks when constructing a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are obviously not equally stable or productive, which means that special attention should be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}).

@@

As explained in the section above, allowing entities using mainly renewables to inject power in the grid is a difficult challenge. Indeed, on the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the resource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Demand side management techniques implemented in the end users' smart meters, can mitigate the consumption peaks and wast less production \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing is believed to be a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

@@

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the returns distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{doi:10.1142/S0219024912500197}.

@@

\begin{equation}
\left\{ \begin{array}{lll}
		\mu_{ij} = \mu_{i} + \mu_{j} \\
		\sigma_{ij} = \sqrt{\sigma_{i}^{2} + \sigma_{j}^{2} + \rho_{ij} \sigma_{i} \sigma_{j} }
\end{array} \right.
\end{equation}
$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants when $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this kind of situations, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalties to the grid. The probability that $ \{i,j\}$ is under-producing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P_{S}^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{S}^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}
where $ erf $ is the error function : $ erf(x) = \dfrac{2}{\sqrt{\pi}}\int_{0}^{x} e^{-t^{2}} dt $.

@@

\begin{equation}
gain(S) = \int_{t_{0}}^{t_{k}} \lambda P_{S}^{CRCT \star} dt = P_{S}^{CRCT \star} \int_{t_{0}}^{t_{k}} \lambda dt
\end{equation} 
(since $ P_{S}^{CRCT \star} $ is supposed to be a constant rate over the contracted period). Using $ gain(S) $ directly as a utility function suffers a major drawback. It is indeed not a concave function of the coalitions' sizes, meaning that coalitions can grow as large as the number of agents allows it, without any counterbalance effects.

@@

\begin{equation}
\mathcal{U}(S) = \left\{ \begin{array}{lll}
							\dfrac{1}{|S|^{\alpha}} \dfrac{ P_{S}^{CRCT \star} }{P^{MAX}},\ if\ S\ is\ valid, \\
							0,\ if\ S\ is\ not\ valid
						 \end{array}
				  \right.
\end{equation}
where parameter $ \alpha $ controls to what extent the size of a coalition impacts its utility, and $ P^{MAX} $ is a normalizing factor. $ P^{MAX} $ can be seen as the maximum production which can be injected in the grid.

@@

In order to get an estimator for $ \alpha $, we solve equation \ref{eq:derivative} in a Gaussian case (as in section \ref{sec:notations}). Furthermore, since considering all the possible interactions between agents is analytically intractable, we use here a mean approximation. Any quantity x that varies over the agent set is thus simplified in its mean value $ \bar{x} $. Solving equation \ref{eq:derivative} for $ \alpha $ in these conditions leads to:

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not to study coalitions stability against player defection which could be done through game theory, nor to redistribute the coalition's utility in terms of individual payoffs. But we aim to design $ \mathcal{U} $ as a measure of how good a given coalition is according to our criteria. In other terms does a given coalitions has a good production to risk ratio.

@@
